---
title: "Day 10 - Temporal Models"
author: "David Carlson"
date: "April 18, 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('~/QRM/inclass/day10temporal/')
library(readstata13)
data = read.dta13('WCRWreplication.dta')
```

# Temporal data analysis

- Repeated measurements over time (but either not TSCS, or no reason to theoretically account for heterogeneity)
- The reading goes over standard temporal analyses, but what if there are not equally temporally spaced observations?
- We will first analyze identity link, modeling the change in polity following a violent or non-violent protest / uprising
- Let's start by looking at the data

```{r violence}
plot(data$politychanget1 ~ data$eyear, type = 'n')
points(data$eyear[as.logical(data$viol)],
       data$politychanget1[as.logical(data$viol)],
       pch = 16, col = 'red', cex = .5)
points(data$eyear[as.logical(data$nonviol)],
       data$politychanget1[as.logical(data$nonviol)],
       pch = 18, col = 'green', cex = .5)
legend('topleft',
       legend = c('Violent', 'Nonviolent'),
       pch = c(16, 18),
       col = c('red', 'green'),
       bty = 'n',
       cex = .5)
```

- Looking at the above plot, it seems that non-violent movements are more associated with increases in polity scores
- But what issues may arise?

```{r linearTemporal}
simpleMod = lm(politychanget1 ~ nonviol, data = data)
summary(simpleMod)
acf(data$politychanget1[order(data$eyear)],
    na.action = na.pass) #not really an issue, but assumes evenly spaced observations
library(car)
qqPlot(simpleMod) #discrete makes it off, but also fat tails
#check for stationarity
#Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test. Here we will test the null hypothesis of trend stationarity (a low p-value will indicate a signal that is not trend stationary, has a unit root) (can also use Augmented Dickey-Fuller Test from readings)
library(tseries)
kpss.test(data$politychanget1[order(data$eyear)],
          null = 'Trend') #low enough to be concerning, and again assumes even spacing
#we could first difference to alleviate the trend, but again, these are not evenly spaced, and the explanatory is binary
#devtools::install_github("andreas50/uts", build_vignettes=TRUE)
library(uts)
# uts(data$politychanget1[order(data$eyear)],
#     as.POSIXct(as.character(data$eyear[order(data$eyear)]),
#                format = '%Y')) #fails, because multiple observations per year
#let's hack it by making them a minute apart
times = c()
for(i in 1:length(unique(data$eyear))) times = c(times, 1:table(data$eyear)[i])
tser = uts(data$politychanget1[order(data$eyear)], 
    as.POSIXct(paste(as.character(data$eyear[order(data$eyear)]),
                 1,
                 times,
                 sep = ':'),
               format = '%Y:%H:%M'))
#now linearly interpolate
newX = sample_values(na.omit(tser),
              as.POSIXct(paste(as.character(rep(min(data$eyear):max(data$eyear), each = max(table(data$eyear)))),
                 1,
                 1:max(table(data$eyear)),
                 sep = ':'),
               format = '%Y:%H:%M'), interpolation  = 'linear')
plot(newX ~ rep(min(data$eyear):max(data$eyear),
                each = max(table(data$eyear))),
     pch = 18)
kpss.test(na.omit(newX),
          null = 'Trend') #now we see a problem
#what about heteroskedasdicity?
#let's do the same for violent or not
tser = uts(data$nonviol[order(data$eyear)], 
    as.POSIXct(paste(as.character(data$eyear[order(data$eyear)]),
                 1,
                 times,
                 sep = ':'),
               format = '%Y:%H:%M'))
#now linearly interpolate
newV = sample_values(na.omit(tser),
              as.POSIXct(paste(as.character(rep(min(data$eyear):max(data$eyear), each = max(table(data$eyear)))),
                 1,
                 1:max(table(data$eyear)),
                 sep = ':'),
               format = '%Y:%H:%M'))
interMod = lm(newX ~ newV)
summary(simpleMod)
summary(interMod) #coefficient does not change much (why?), but SEs decrease (why?)
acf(newX, na.action = na.pass) #now we clearly see there is a problem
#what do we do?
#the explanatory is binary, so we cannot first difference or lag or anything typical
#we can adjust the standard errors (Newey-West)
library(sandwich)
library(lmtest)
coeftest(interMod,
         vcov. = NeweyWest(interMod)) #we see an increase in the SE
data2 = na.omit(data[, c('politychanget1',
                         'viol',
                         'eyear')])
simpleMod2 = lm(politychanget1 ~ viol, data2)
coeftest(simpleMod2,
         vcov. = NeweyWest(simpleMod2,
                           order.by = ~data2$eyear))
#the above is improper, because again assumes evenly spaced observations
#Newey-West also makes strong parametric assumptions
#this is when we can turn to GPs (more details next week)
library(kernlab)
modGP = gausspr(politychanget1 ~ viol + eyear,
        data = data2,
        kernel = 'rbfdot',
        type = 'regression',
        variance.model = T) #we use dot product for the non-stationarity, but here we use sep exp
#when we code our own model, we can specify different inputs to mean and kernel, and make inferences on correlations
#for now we just plot to make inferences
xtestViol = data.frame('viol' = 1,
                       'eyear' = min(data2$eyear):max(data2$eyear))
xtestNonViol = data.frame('viol' = 0,
                          'eyear' = min(data2$eyear):max(data2$eyear))
ysViol = predict(modGP, xtestViol)
ysViolLower = ysViol - 1.96*predict(modGP,
                                    xtestViol,
                                    type = 'sdeviation')
ysViolUpper = ysViol + 1.96*predict(modGP,
                                    xtestViol,
                                    type = 'sdeviation')
ysNonViol = predict(modGP, xtestNonViol)
ysNonViolLower = ysNonViol - 1.96*predict(modGP,
                                    xtestNonViol,
                                    type = 'sdeviation')
ysNonViolUpper = ysNonViol + 1.96*predict(modGP,
                                    xtestNonViol,
                                    type = 'sdeviation')

plot(data$politychanget1 ~ data$eyear, type = 'n')
points(data$eyear[as.logical(data$viol)],
       data$politychanget1[as.logical(data$viol)],
       pch = 16, col = 'red', cex = .5)
points(data$eyear[as.logical(data$nonviol)],
       data$politychanget1[as.logical(data$nonviol)],
       pch = 18, col = 'green', cex = .5)
lines(min(data2$eyear):max(data2$eyear),
      ysViol, col = 'red')
lines(min(data2$eyear):max(data2$eyear),
      ysViolUpper, col = 'red', lty = 2)
lines(min(data2$eyear):max(data2$eyear),
      ysViolLower, col = 'red', lty = 2)
lines(min(data2$eyear):max(data2$eyear),
      ysNonViol, col = 'green')
lines(min(data2$eyear):max(data2$eyear),
      ysNonViolUpper, col = 'green', lty = 2)
lines(min(data2$eyear):max(data2$eyear),
      ysNonViolLower, col = 'green', lty = 2)

legend('topleft',
       legend = c('Violent', 'Nonviolent'),
       pch = c(16, 18),
       col = c('red', 'green'),
       bty = 'n',
       cex = .5)
#we see that once temporal trends are accounted for, we cannot reject the null (but of course we are not controlling for anything, so take it with a grain of salt)
```

# Estimation of Growth Rates

- There are many types of growth models
- They generally fall into two categories: Population dynamics in demography, and economic growth
- We will model the growth rate of protest movements, violent and non-violent, over time, and compare the growth rates

```{r growth}
library(growthrates)
dataViol = data[as.logical(data$viol), ]
dataNonViol = data[!as.logical(data$viol), ]
#linear growth model
growthLinViol = fit_easylinear(as.numeric(names(table(dataViol$eyear))),
                               as.numeric(table(dataViol$eyear)))
growthLinNonViol = fit_easylinear(as.numeric(names(table(dataNonViol$eyear))),
                               as.numeric(table(dataNonViol$eyear)))
summary(growthLinViol)
summary(growthLinNonViol) #growing much faster

#Nonparametric smoothing splines
#Smoothing splines are a quick method to estimate maximum growth. The method is called nonparametric, because the growth rate is directly estimated from the smoothed data without being restricted to a specific model formula
growthNPViol = fit_spline(as.numeric(names(table(dataViol$eyear))),
                               as.numeric(table(dataViol$eyear)))
growthNPNonViol = fit_spline(as.numeric(names(table(dataNonViol$eyear))),
                               as.numeric(table(dataNonViol$eyear)))
par(mfrow = c(1, 2))
plot(growthNPViol)
plot(growthNPNonViol)
#we see that non-violent growth is estimated to be consistent, while violent growth is estimated to be quite wavy
```




